Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-04 doc/invt.tex
invt.c: correct mu for pbc
optimal t_0 = 2/alpha_0

Since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix $\mathbf w$
with identical eigenvalues
is essentially a multiple of the identity matrix.
%
Thus, if no eigenvalue $\lambda_k$ is zero,
then in terms of asymptotic convergence,
the single-bin scheme adopted in the WL algorithm
is most efficient.
%

Note that
since we only optimized the asymptotic error,
the inverse-time schedule is optimal
only in the long time limit,
but not necessarily so for a finite-length simulation.
%
For example, consider a system that is initially
prepared under a fixed $\alpha_0$
at the beginning of the schedule $\alpha(t)$.
%
Then for shorter times, $T \approx 0$,
the error, dominated by the residual error
given by Eq. \eqref{eq:error_res1},
is clearly smaller if some eigenvalues
are less than $1.0$ (the single-bin case).
%
%In practice, real simulations
%may have long achieved the desired precision
%before reaching the asymptotic regime,
%especially if the updating matrix
%has some $\lambda_k$'s close to $0$.

\subsubsection{\label{sec:optscheme}
Cardinal updating schemes}

We can also generalize
the single-bin scheme to a class of
optimal updating schemes
that allow zero eigenvalues.
%
For example,
if we can assume that the PMF is sufficiently smooth
such that the error modes can be limited to $k \le K$.
%
Then we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
Eq. \eqref{eq:error_singlebin} is modified to
$$
  E
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
$$
The optimal schedule of these schemes
remains to be the inverse-time schedule,
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

For a periodic variable, if we demand
%
$$
\begin{aligned}
&
\lambda_1 = \cdots = \lambda_{K+1}
= \lambda_{n-K+1} = \cdots = \lambda_n = 1,
\\
&
\lambda_{K+2} = \cdots = \lambda_{n-K} = 0,
\end{aligned}
$$
for $K \ge 1$, then by using
Eq. \eqref{eq:mu_from_lambda_pbc},
we get
\begin{equation}
  \mu_l
  =
  \frac{
    \sin
    \dfrac{ (2 K + 1) \, l \, \pi }
         {              n        }
  }
  {
    n \, \sin \dfrac{ l \, \pi } { n }
  }
  .
  \label{eq:mu_cardinal_pbc}
\end{equation}
\note{Derivation.
$$
\begin{aligned}
\mu_l
&=
\frac 1 n \sum_{k = 0}^{n-1} \lambda_{k+1} \cos \frac{ k \, l \, 2 \, \pi } { n }
\\
&=
\frac{1}{n}
\left(
  1 +
  \sum_{k=1}^K
  \cos \frac { k \, l \, 2 \, \pi } { n }
  +
  \sum_{k=n-K}^{n-1}
  \cos \frac { k \, l \, 2 \, \pi } { n }
\right)
\\
&=
\frac 1 n
\sum_{k=-K}^K
\cos \frac { k \, l \, 2 \, \pi } { n }
=
  \frac{
    \sin
    \dfrac{ (2 K + 1) \, l \, \pi }
         {              n        }
  }
  {
    n \, \sin \dfrac{ l \, \pi } { n }
  }
.
\end{aligned}
$$
}

For a non-periodic variable, we demand
$$
\lambda_1 = \cdots = \lambda_{K+1} = 1,
\qquad
\lambda_{K+2} = \cdots = \lambda_n = 0.
$$
Then Eq. \eqref{eq:lambdan} gives
$\lambda_n = (-1)^{n+K-1}$,
and from Eq. \eqref{eq:mu_from_lambda_refl},
we have
\begin{equation}
  \mu_l
  =
  \frac{1}{2 \, n}
  \left[
    (-1)^{n+K+l-1}
    +
    \frac{
      \sin
      \dfrac{ (2 K + 1) \, l \, \pi }
           {         2 \, n        }
    }
    {
      n \, \sin \dfrac{ l \, \pi } { 2 \, n }
    }
  \right]
  .
\end{equation}
\note{Derivation.
$$
\begin{aligned}
  \lambda_{n+1}
  &=
  (-1)^{n-1}
  \left[
    \lambda_1
    + 2 (-\lambda_2 + \lambda_3 - \cdots + (-1)^K \lambda_{K+1})
  \right]
  \\
  &=
  \begin{dcases}
    (-1)^{n-1} & K \mathrm{\; even,} \\
    (-1)^n     & K \mathrm{\; odd.}
  \end{dcases}
\end{aligned}
$$
So
$$
\begin{aligned}
  \lambda_l
  &=
  \frac{1}{2\,n}
  \left[
    1 +
    2 \sum_{k=2}^{K+1}
    \cos \frac { (k - 1) \, l \pi } { n }
    +
    (-1)^{n+K-1} (-1)^l
  \right]
  \\
  &=
  \frac{1}{2\,n}
  \left[
    (-1)^{n+K+l-1}
    +
    \sum_{k=-K}^{K}
    \cos \frac { k \, l \pi } { n }
  \right]
  .
\end{aligned}
$$

@@

>>>>

Since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix $\mathbf w$
with identical eigenvalues
is essentially a multiple of the identity matrix.
%
Thus, if no eigenvalue $\lambda_k$ is zero,
then in terms of asymptotic convergence,
the single-bin scheme adopted in the WL algorithm
is most efficient.

\subsubsection{\label{sec:optscheme}
Brick-wall band-pass updating schemes}

We can also generalize
the single-bin scheme to a class of
\emph{brick-wall band-pass} updating schemes
that allow zero eigenvalues.
%
For example,
if we can assume that the PMF is sufficiently smooth
such that the error modes can be limited to $k \le K$.
%
Then we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
Eq. \eqref{eq:error_singlebin} is changed to
%
\begin{equation}
  E_A
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time schedule,
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

For a periodic variable, if we demand
%
$$
\begin{aligned}
&
\lambda_1 = \cdots = \lambda_{K+1}
= \lambda_{n-K+1} = \cdots = \lambda_n = 1,
\\
&
\lambda_{K+2} = \cdots = \lambda_{n-K} = 0,
\end{aligned}
$$
for $K \ge 1$, then by using
Eq. \eqref{eq:mu_from_lambda_pbc},
we get
\begin{equation}
  \mu_l
  =
  \frac{
    \sin
    \dfrac{ (2 K + 1) \, l \, \pi }
         {              n        }
  }
  {
    n \, \sin \dfrac{ l \, \pi } { n }
  }
  .
  \label{eq:mu_sinc_pbc}
\end{equation}
\note{Derivation.
$$
\begin{aligned}
\mu_l
&=
\frac 1 n \sum_{k = 0}^{n-1} \lambda_{k+1} \cos \frac{ k \, l \, 2 \, \pi } { n }
\\
&=
\frac{1}{n}
\left(
  1 +
  \sum_{k=1}^K
  \cos \frac { k \, l \, 2 \, \pi } { n }
  +
  \sum_{k=n-K}^{n-1}
  \cos \frac { k \, l \, 2 \, \pi } { n }
\right)
\\
&=
\frac 1 n
\sum_{k=-K}^K
\cos \frac { k \, l \, 2 \, \pi } { n }
=
  \frac{
    \sin
    \dfrac{ (2 K + 1) \, l \, \pi }
         {              n        }
  }
  {
    n \, \sin \dfrac{ l \, \pi } { n }
  }
.
\end{aligned}
$$
}

For a non-periodic variable, we demand
$$
\lambda_1 = \cdots = \lambda_{K+1} = 1,
\qquad
\lambda_{K+2} = \cdots = \lambda_n = 0.
$$
Then Eq. \eqref{eq:lambdan} gives
$\lambda_n = (-1)^{n+K-1}$,
and from Eq. \eqref{eq:mu_from_lambda_refl},
we have
\begin{equation}
  \mu_l
  =
  \frac{1}{2 \, n}
  \left[
    (-1)^{n+K+l-1}
    +
    \frac{
      \sin
      \dfrac{ (2 K + 1) \, l \, \pi }
           {         2 \, n        }
    }
    {
      n \, \sin \dfrac{ l \, \pi } { 2 \, n }
    }
  \right]
  .
\label{eq:mu_sinc_refl}
\end{equation}
\note{Derivation.
$$
\begin{aligned}
  \lambda_{n+1}
  &=
  (-1)^{n-1}
  \left[
    \lambda_1
    + 2 (-\lambda_2 + \lambda_3 - \cdots + (-1)^K \lambda_{K+1})
  \right]
  \\
  &=
  \begin{dcases}
    (-1)^{n-1} & K \mathrm{\; even,} \\
    (-1)^n     & K \mathrm{\; odd.}
  \end{dcases}
\end{aligned}
$$
So
$$
\begin{aligned}
  \lambda_l
  &=
  \frac{1}{2\,n}
  \left[
    1 +
    2 \sum_{k=2}^{K+1}
    \cos \frac { (k - 1) \, l \pi } { n }
    +
    (-1)^{n+K-1} (-1)^l
  \right]
  \\
  &=
  \frac{1}{2\,n}
  \left[
    (-1)^{n+K+l-1}
    +
    \sum_{k=-K}^{K}
    \cos \frac { k \, l \pi } { n }
  \right]
  .
\end{aligned}
$$

@@

\subsubsection{\label{sec:finlencorr}
Finite-length correction}

In the above, we have been focusing on
the asymptotic error.
%
For a finite-length simulation,
we should include the residual part,
Eq. \eqref{eq:error_res}
and minimize the total error.
%
For the brick-wall band-pass updating schemes,
this allows to determine the optimal value
of the free parameter, $t_0$.

Assuming the system is initially equilibrated
under a constant updating magnitude, $\alpha_0$,
we may use Eq. \eqref{eq:error_res1} with
$$
q(T)
=
\int \frac{ dt } { t + t_0 }
=
\ln \frac{ T + t_0 } { t_0 }.
$$
%
Thus, from Eqs. \eqref{eq:error_split}
and \eqref{eq:error_asym_sinc}, we have
%
\begin{equation}
  E
  =
  E_R + E_A
  =
  \frac { \frac 1 2 \alpha_0 \, t_0^2 + T }
        {          ( T + t_0 )^2          }
  \sum_{ k = 2 }^K \Gamma_k
  .
\notag
%\label{eq:error_sinc0}
\end{equation}
%
The optimal error is reached at
%
\begin{equation}
  t_0
  =
  \frac{     2    }
       { \alpha_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
and at this point
%
\begin{equation}
  E
  =
  \frac{        1         }
       { T + 2 / \alpha_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
This means that at the beginning of the schedule,
the updating magnitude is
$\alpha(0^+) = \alpha_0 / 2$,
half of the previous equilibrium value.
%
Interestingly, this factor of $1/2$
happens to coincide with the suggested
reduction factor of updating magnitude
in the original WL algorithm\cite{wang2001, wang2001pre}
during stage transitions.

%
%For example, consider a system that is initially
%prepared under a fixed $\alpha_0$
%at the beginning of the schedule $\alpha(t)$.
%%
%Then for shorter times, $T \approx 0$,
%the error, dominated by the residual error
%given by Eq. \eqref{eq:error_res1},
%is clearly smaller if some eigenvalues
%are less than $1.0$ (the single-bin case).
%
%In practice, real simulations
%may have long achieved the desired precision
%before reaching the asymptotic regime,
%especially if the updating matrix
%has some $\lambda_k$'s close to $0$.
